Imputation of Incomplete Multilevel Data with \pkg{mice}

### 2022-02-04 Manuscript/Manuscript.Rmd
Add analysis model popularity data

The variability due to clustering is often measured by means of the intra-class coefficient (ICC). The ICC can be seen as the percentage of variance that can be attributed to the cluster-level, where a high ICC would indicate that a lot of variability is due to the cluster structure. Multilevel models typically accommodate for this variability by including separate intercepts for each cluster. Such fixed effects relieve the restriction imposed by single-level models: equal group means across clusters. Additionally, models may include random effects of predictors across clusters, and random error terms (heterogeneous residual error variances; see e.g. @hox17 and @jong21). There are many names for models that take clustering into account. Some popular examples are 'multilevel models', 'hierarchical models', 'mixed effect models' and 'random effect models'.

@@

This papers serves as a tutorial for imputing incomplete multilevel data with \pkg{mice}. We provide practical guidelines and code snippets for different missing data situations. For reasons of brevity, we focus on imputation by chained equations exclusively (although the alternative, joint modeling imputation for multilevel data or \pkg{jomo} @jomo, has been implemented in \pkg{mice} as well). Other useful resources for the analysis of incomplete multilevel data include the \proglang{R} packages \pkg{mitml}, \pkg{miceadds}, and \pkg{mdmb}, and the empirical work by @audi18 and @grun18. Note that this tutorial paper assumes a basic level of knowledge on multilevel models. We're providing an overview of implementations. It's up-to the reader to decide which multilevel strategy suits their data. We won't go into detail for the different methods (and equations). This paper is just a software tutorial, so we'll keep it practical.

@@

- `mice::popmis` (simulated data on perceived popularity, $n = 199$ pupils across $N = 10$ classes, with univariate MAR missingness);
- `GJRM::hiv` (simulated data on HIV diagnoses, $n = 6,416$ patients across $N = 9$ regions, with univariate MNAR missingness)
- `metamisc::impact` (empirical data on traumatic brain injuries, $n = 11,022$ patients across $N = 15$ studies, without `NA`s).

@@

# load data 
popmis <- mice::popmis %>% mutate(
  class = factor(school), 
  .keep = "unused")#readRDS("../Data/popNCR2.RDS") %>% 
  #filter(class %in% clusters) %>% 
  #mutate(class = as.factor(class))
popcomp <- foreign::read.spss("../Data/popular2.sav", to.data.frame = TRUE)
```

In this section we'll go over the different steps involved with imputing incomplete multilevel data. The data we're using is the `popmis` dataset from the `mice` package. This is a simulated dataset with pupils clustered in classes, where the number of units $n = 2000$, and the number of clusters $N = 100$, on 7 variables:

  - `pupil`	Pupil number within class,
  - `class`	Class number,
  - `extrav`	Pupil extraversion,
  - `sex`	Pupil gender,
  - `texp`	Teacher experience (years),
  - `popular`	Pupil popularity,
  - `teachpop`	Teacher popularity.

The analysis model corresponding to this dataset is multilevel regression with `popular` as the outcome variable, `sex` and `extrav` as unit-level predictors, and `texp` and `teachpop` as cluster-level predictors. Since the data is simulated and the missingness is induced, we can compare our imputations to the true data generating mechanism.

The data is created in such a way that the clustering variable `class` explains quite some variance in the outcome variable `popular`. We express this using the intra-class correlation,  ICC $=$ `r round(icc(popular ~ as.factor(class), data = popcomp), 2)`. This tells us that it would be wise to use multilevel modeling.

@@

@@

The missingness is univariate and sporadic, which is illustrated in the missing data pattern in Figure \ref{fig:pop_pat}. The ICC in the incomplete data is `r round(icc(popular ~ as.factor(class), data = na.omit(popmis)), 2)`.

To develop the best imputation model for the incomplete variable `popular`, we need to know whether the missingness depends on the observed values of other variables. We'll highlight only one other variable to illustrate, but ideally one would inspect all relations. The questions we'll ask are: 'Does the missing data of pupil popularity (`popular`) depend on observed teacher popularity (`teachpop`)?'. This can be evaluated statistically, but visual inspection usually suffices. We'll make a histogram of `teachpop` separately for the pupils with known popularity and missing popularity.

In Figure \ref{fig:pop_dist} we see that **[update this part]** the distribution for the missing `popular` is further to the right than the distribution for observed `popular`. This would indicate a right-tailed MAR missingness. (In fact, this is exactly what happens, because the missingness in these data was created manually.) We've made it observable by examining the relations between the missingness in popular and the observed data in `teachpop`. 
<!-- There is also a dependency between the missingness in teacher popularity and pupil popularity. The relation seems to be right-tailed as well. -->

```{r pop_dist, fig.cap = "Conditional distributions in the popularity data", echo=FALSE, message=FALSE, warning=FALSE}
# popmis %>% 
#   filter(as.numeric(class) < 5) %>% 
#   ggplot(aes(x = teachpop, color = class)) +
#     geom_density() +
#     facet_wrap(~factor(is.na(popular), labels = c("popular missing", "popular observed")), ncol = 1) +
#     # scale_color_manual(values = plot_col, name = "popular") +
#     theme_classic() # +
#     # theme(legend.position = "bottom")

plot_conditional(popmis, x = "teachpop", z = "popular", cluster = "class")
# TODO: check if up-side-down plot works
# TODO: check smoothing in geom_density function and make it the inverse of the sample size
# TODO: add functions to mice
# TODO: add title and informative legend
# TODO: add facets for some clusters
# or add propensity score distribution `is.na(popular) ~ .`
# TODO: make it average of cluster dens, not marginal
# alternatively add a quartile line around the density with geom_ribbon

@@

We'll use predictive mean matching to impute the continuous variables and logistic regression to impute the binary variable `sex`. We do not use the observation identifier `pupil` or cluster identifier `class` as predictors to impute other variables.

@@

# extract predictor matrix and adjust
pred <- ini$pred
pred[, c("class", "pupil")] <- 0

@@

# compare ICCs before and after imputation
ICCs <- data.frame(
  vars = c("popular", "teachpop", "texp"), 
  CCA = c(icc(popular ~ class, popmis), 
               icc(teachpop ~ class, popmis),
               icc(texp ~ class, popmis)), 
  ignored = c(icc(popular ~ class, complete(imp_ignored)), 
              icc(teachpop ~ class, complete(imp_ignored)), 
              icc(texp ~ class, complete(imp_ignored)))
  )
ICCs

@@

As the original ICCs show, 100% of the variance in `texp` can be attributed to the clustering variable `class`. This tells us that the multilevel structure of the data should be taken into account. If we don't, we'll end up with incorrect imputations, biasing the effect of the clusters towards zero.

We can also observe that the teacher experience increases slightly after imputation. This is due to the MNAR missingness in `texp`. Higher values for `texp` have a larger probability to be missing. This may not a problem, however, if at least one pupil in each class has teacher experience recorded, we can deductively impute the correct (i.e. true) value for every pupil in the class.

@@

We'll now use `class` as a predictor to impute all other variables. This is still not recommended practice, since it only works under certain circumstances and results may be biased [@drec15; @ende16]. But at least, it includes some multilevel aspect. This method is also called 'fixed cluster imputation', and uses N-1 indicator variables representing allocation of N clusters as a fixed factor in the model [@reit06; @ende2016]. Colloquially, this is 'multilevel imputation for dummies'.

@@

```{r pop_predictor, message=FALSE, warning=FALSE}
# adjust the predictor matrix
pred <- ini$pred 
pred[, "pupil"] <- 0

@@

# compare ICCs before and after imputation
ICCs <- ICCs %>% mutate(
           predictor = c(icc(popular ~ class, complete(imp_predictor)), 
                        icc(teachpop ~ class, complete(imp_predictor)), 
                        icc(texp ~ class, complete(imp_predictor)))
           )
ICCs
```

@@

```{r pop_norm}
pred <- ini$pred
pred["popular", ] <- c(0, -2, 2, 2, 2, 0, 2) 
#-2 for the cluster variable, 2 for random effects
meth <- ini$meth
meth <- c("", "", "", "", "", "2l.norm", "")
imp_norm_2l <-
  mice(
    popmis %>% mutate(class = as.integer(class)),
    pred = pred,
    meth = meth,
    maxit = 1,
    print = FALSE
  )
```

@@

ICCs <- ICCs %>% mutate(
           norm = c(icc(popular ~ as.factor(class), complete(imp_norm_2l)), 
                    icc(teachpop ~ as.factor(class), complete(imp_norm_2l)), 
                    icc(texp ~ as.factor(class), complete(imp_norm_2l)))
           )
ICCs
```

@@

```{r pop_pan}
pred["popular", ] <- c(0, -2, 2, 2, 1, 0, 2)
meth <- c("", "", "", "", "", "2l.pan", "")
imp_pan_2l <-
  mice(
    popmis %>% mutate(class = as.integer(class)),
    pred = pred,
    meth = meth,
    maxit = 1,
    print = FALSE
  )

@@

```{r pop_pan_eval, echo=FALSE}
# plot(imp_pan)
plot_imps(imp_pan_2l, x = "popular", type = "stripplot")
ICCs <- ICCs %>% mutate(
           pan = c(icc(popular ~ as.factor(class), complete(imp_pan_2l)), 
                    icc(teachpop ~ as.factor(class), complete(imp_pan_2l)), 
                    icc(texp ~ as.factor(class), complete(imp_pan_2l)))
           )
ICCs
```

>>>>

The variability due to clustering is often measured by means of the intraclass coefficient (ICC). The ICC can be seen as the percentage of variance that can be attributed to the cluster-level, where a high ICC would indicate that a lot of variability is due to the cluster structure. Multilevel models typically accommodate for this variability by including separate intercepts for each cluster. Such fixed effects relieve the restriction imposed by single-level models: equal group means across clusters. Additionally, models may include random effects of predictors across clusters, and random error terms (heterogeneous residual error variances; see e.g. @hox17 and @jong21). There are many names for models that take clustering into account. Some popular examples are 'multilevel models', 'hierarchical models', 'mixed effect models' and 'random effect models'.

@@

This papers serves as a tutorial for imputing incomplete multilevel data with \pkg{mice}. We provide practical guidelines and code snippets for different missing data situations. For reasons of brevity, we focus on imputation by chained equations exclusively (although the alternative, joint modeling imputation for multilevel data or \pkg{jomo} @jomo, has been implemented in \pkg{mice} as well). Other useful resources for the analysis of incomplete multilevel data include the \proglang{R} packages \pkg{mitml}, \pkg{miceadds}, and \pkg{mdmb}, and the empirical work by @audi18 and @grun18. Note that this tutorial paper assumes a basic level of knowledge on multilevel models. We're providing an overview of implementations. It's up-to the reader to decide which multilevel strategy suits their data. We won't go into detail for the different methods (and equations). This paper is just a software tutorial, so we'll keep it practical. Assumed knowledge also includes the use of the 'piping operator', `%>%`, adopted from the `magrittr` package and avidly used among `tidy` workflows, and the `lme4` notation for multilevel models.

@@

- `mice::popmis` (simulated data on perceived popularity, $n = 2,000$ pupils across $N = 100$ schools, with univariate MAR missingness);
- `GJRM::hiv` (simulated data on HIV diagnoses, $n = 6,416$ patients across $N = 9$ regions, with univariate MNAR missingness)
- `metamisc::impact` (empirical data on traumatic brain injuries, $n = 11,022$ patients across $N = 15$ studies, without `NA`s).

@@

# load data 
# popmis <- mice::popmis 
#readRDS("../Data/popNCR2.RDS") %>% 
  #filter(class %in% clusters) %>% 
  #mutate(class = as.factor(class))
popcomp <- foreign::read.spss("../Data/popular.sav", to.data.frame = TRUE) %>% 
  mutate(school = SCHOOL,
         popular = as.numeric(POPULAR),
         sex = SEX,
         texp = TEXP,
         .keep = "none")
comp_est <- lme4::lmer(popular ~ 1 + sex + texp + (1 + sex | school), popcomp) %>% broom.mixed::tidy()
```

In this section we'll go over the different steps involved with imputing incomplete multilevel data. The data we're using is the `popmis` dataset from the `mice` package. This is a simulated dataset with pupils ($n = 2000$) clustered within schools ($N = 100$). In this tutorial we'll use the following variables:

  - `school`,     school identification number (clustering variable);
  - `popular`,    pupil popularity (self-rating between 0 and 10; unit-level);
  - `sex`,	      pupil sex (0=boy, 1=girl; unit-level);
  - `texp`,	      teacher experience (in years; cluster-level).

The analysis model corresponding to this dataset is multilevel regression with random intercepts, random slopes and a cross-level interaction. The outcome variable is `popular`, which is predicted from the unit-level variable `sex` and the cluster-level variable `texp`. The regression equation and `lme4` notation for this model are

$$
\text{popular}_{ij} =
\gamma_{00} + 
\gamma_{10} \text{ sex}_{ij} + 
\gamma_{01} \text{ texp}_{j} + 
\gamma_{11} \text{ texp}_{j} \times \text{sex}_{ij} + 
u_{0j} + 
u_{1j} \text{ sex}_{ij} + 
e_{ij} \\
$$

$$
\texttt{popular} \sim  \texttt{1 + sex + texp + sex:texp + (1 + sex | school)}
$$

<!-- Intercept-only model: lmer(formula = popular ~ 1 + (1|class)) -->
<!-- First level predictors: lmer(formula = popular ~ 1 + sex + (1|class)) -->
<!-- First and second level predictors: lmer(popular ~ 1 + sex + texp + (1 | class)) -->
<!-- First and second level predictors with random slopes: lmer(formula = popular ~ 1 + sex + texp + (1 + sex | class) -->
<!-- First and second level predictors with random slopes and cross-level interaction: lmer(formula = popular ~ 1 + sex + texp + sex:texp + (1 + sex | class)) -->

Since the data is simulated and the missingness is induced, we can compare our inferences after imputation to the true complete data. The data is created in such a way that the clustering variable `school` explains quite some variance in the outcome variable `popular`. We express this using the intraclass correlation,  ICC $=$ `r round(icc(popular ~ as.factor(school), data = popcomp), 2)`. We'll evaluate the ICC after each missing data strategy, and compare the estimated fixed effect of `sex` on `popular` $\gamma_{10} = 0.843$.

@@

Load the data into the environment and select the relevant variables:

```{r}
popmis <- mice::popmis[, c("school", "popular", "sex", "texp")] 
```

@@

The missingness is univariate and sporadic, which is illustrated in the missing data pattern in Figure \ref{fig:pop_pat}. The ICC in the incomplete data is `r round(icc(popular ~ as.factor(school), data = na.omit(popmis)), 2)`.

To develop the best imputation model for the incomplete variable `popular`, we need to know whether the missingness depends on the observed values of other variables. We'll highlight only one other variable to illustrate, but ideally one would inspect all relations. The questions we'll ask are: 'Does the missing data of pupil popularity (`popular`) depend on observed teacher popularity (`texp`)?'. This can be evaluated statistically, but visual inspection usually suffices. We'll make a histogram of `texp` separately for the pupils with known popularity and missing popularity.

In Figure \ref{fig:pop_dist} we see that **[update this part]** the distribution for the missing `popular` is further to the right than the distribution for observed `popular`. This would indicate a right-tailed MAR missingness. (In fact, this is exactly what happens, because the missingness in these data was created manually.) We've made it observable by examining the relations between the missingness in popular and the observed data in `texp`. 
<!-- There is also a dependency between the missingness in teacher popularity and pupil popularity. The relation seems to be right-tailed as well. -->

```{r pop_dist, fig.cap = "Conditional distributions in the popularity data", echo=FALSE, message=FALSE, warning=FALSE}
# popmis %>% 
#   filter(as.numeric(school) < 5) %>% 
#   ggplot(aes(x = teachpop, color = school)) +
#     geom_density() +
#     facet_wrap(~factor(is.na(popular), labels = c("popular missing", "popular observed")), ncol = 1) +
#     # scale_color_manual(values = plot_col, name = "popular") +
#     theme_classic() # +
#     # theme(legend.position = "bottom")

plot_conditional(popmis, x = "texp", z = "popular", cluster = "school")
# TODO: check if up-side-down plot works
# TODO: check smoothing in geom_density function and make it the inverse of the sample size
# TODO: add functions to mice
# TODO: add title and informative legend
# TODO: add facets for some clusters
# or add propensity score distribution `is.na(popular) ~ .`
# TODO: make it average of cluster dens, not marginal
# alternatively add a quartile line around the density with geom_ribbon

@@

We'll use predictive mean matching to impute the continuous variables and logistic regression to impute the binary variable `sex`. We do not use the observation identifier `pupil` or cluster identifier `school` as predictors to impute other variables.

@@

# extract predictor matrix and adjust
pred <- ini$pred
pred[, c("school")] <- 0

@@

# # compare ICCs before and after imputation
# ICCs <- data.frame(
#   vars = c("popular", "teachpop", "texp"), 
#   CCA = c(icc(popular ~ school, popmis), 
#                icc(teachpop ~ school, popmis),
#                icc(texp ~ school, popmis)), 
#   ignored = c(icc(popular ~ school, complete(imp_ignored)), 
#               icc(teachpop ~ school, complete(imp_ignored)), 
#               icc(texp ~ school, complete(imp_ignored)))
#   )
# ICCs

@@

As the original ICCs show, 100% of the variance in `texp` can be attributed to the clustering variable `school`. This tells us that the multilevel structure of the data should be taken into account. If we don't, we'll end up with incorrect imputations, biasing the effect of the clusters towards zero.

We can also observe that the teacher experience increases slightly after imputation. This is due to the MNAR missingness in `texp`. Higher values for `texp` have a larger probability to be missing. This may not a problem, however, if at least one pupil in each school has teacher experience recorded, we can deductively impute the correct (i.e. true) value for every pupil in the school.

@@

We'll now use `school` as a predictor to impute all other variables. This is still not recommended practice, since it only works under certain circumstances and results may be biased [@drec15; @ende16]. But at least, it includes some multilevel aspect. This method is also called 'fixed cluster imputation', and uses N-1 indicator variables representing allocation of N clusters as a fixed factor in the model [@reit06; @ende2016]. Colloquially, this is 'multilevel imputation for dummies'.

@@

```{r pop_predictor, message=FALSE, warning=FALSE}
# adjust the predictor matrix
pred <- ini$pred 
# pred[, "pupil"] <- 0

@@

# # compare ICCs before and after imputation
# ICCs <- ICCs %>% mutate(
#            predictor = c(icc(popular ~ school, complete(imp_predictor)), 
#                         icc(teachpop ~ school, complete(imp_predictor)), 
#                         icc(texp ~ school, complete(imp_predictor)))
#            )
# ICCs
```

@@

```{r pop_norm}
pred <- ini$pred
pred["popular", ] <- c(-2, 2, 2, 2) 
#-2 for the cluster variable, 2 for random effects
meth <- ini$meth
meth <- c("", "2l.norm", "",  "")
imp_norm_2l <-
  mice(
    popmis %>% mutate(school = as.integer(school)),
    pred = pred,
    meth = meth,
    maxit = 1,
    print = FALSE
  )
```

@@

# ICCs <- ICCs %>% mutate(
#            norm = c(icc(popular ~ as.factor(school), complete(imp_norm_2l)), 
#                     icc(teachpop ~ as.factor(school), complete(imp_norm_2l)), 
#                     icc(texp ~ as.factor(school), complete(imp_norm_2l)))
#            )
# ICCs
```

@@

```{r pop_pan}
pred["popular", ] <- c(-2, 2, 1, 2)
meth <- c("", "2l.pan", "", "")
imp_pan_2l <-
  mice(
    popmis %>% mutate(school = as.integer(school)),
    pred = pred,
    meth = meth,
    maxit = 1,
    print = FALSE
  )

@@

```{r pop_pan_eval, echo=FALSE}
# plot(imp_pan)
plot_imps(imp_pan_2l, x = "popular", type = "stripplot")
# ICCs <- ICCs %>% mutate(
#            pan = c(icc(popular ~ as.factor(school), complete(imp_pan_2l)), 
#                     icc(teachpop ~ as.factor(school), complete(imp_pan_2l)), 
#                     icc(texp ~ as.factor(school), complete(imp_pan_2l)))
#            )
# ICCs
```
